Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-11-01 src/tex/ms.tex
[showyourwork] overleaf sync

% Filler text
\usepackage{blindtext}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{showyourwork}
\usepackage[autostyle]{csquotes}
\usepackage{amssymb}
\let\tablenum\relaxgit 
\usepackage{siunitx}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\arcsinh}{arcsinh}
\newcommand{\chandra}{\textit{Chandra}~}
\newcommand{\xmm}{\textit{XMM}~}
\newcommand{\fermi}{\textit{Fermi}-LAT~}
\newcommand{\jolideco}{\textit{Jolideco}~}
 \newcommand{\aposteriori}{a~posteriori~}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\xray}{X-ray\xspace}
\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}\PackageWarning{TODO:}{#1!}}

@@

    \section{Introduction}
    The resolution of any physical or astronomical imaging process is inherently limited by the instrument or telescope used. Furthermore, the quality of the resulting image is affected by background or instrumental measurement noise and non-uniform exposure. This is especially true in the low signal-to-noise regime, where images are affected by Poisson noise. This includes images taken by \xray and \gammaray~telescopes but also medical images obtained with methods such as positron emission tomography or single-photon emission computed tomography.

    To maximise the scientific use of available data it is desirable to correct for the instrumental imprint on the data as well as reduce its noise level. The process of improving the resolution of an image in post-processing is often called \enquote{deblurring} or \enquote{deconvolution}, as it aims to invert the effect of the data being degraded by a forward convolution through the point spread function (PSF) of the instrument. In general this class of problems is categorised as \enquote{inverse problems}. If there is prior information available on the shape of the PSF, we typically refer to the process more precisely as \enquote{unblind deconvolution}. In case of astronomical observations the PSF can for example be estimated empirically from observation of known point sources, such as distant \as{I would say stars instead of AGN - in optical and even in X-rays telescopes AGN are resolved, so not point sources} activate galactic nuclei or a model of the PSF can obtained from simulation of the telescope or instrument.

\as{You may want to add a short para or just a sentance saying that X-ray and gamma-ray images are the Poisson type images containing low number of counts in each image pixel etc.. - this is to link to the next subsection which is specific to Poisson noise?}

@@

    The standard baseline method for image deconvolution in the presence of Poisson noise was proposed independently by \cite{Richardson1972} and \cite{Lucy1974} (we will further refer to this method as \enquote{\textit{RL}}). The method involves estimating the reconstructed image by an iterative procedure that converges to a maximum likelihood solution of the problem.

@@

    \cite{Selig2015} and \cite{Pumpe2018} approached the problem of deconvolution of Poisson noise data implicitly using the framework of information field theory. Based on physical prior assumptions on the spatial distribution of point sources and local correlation structures of diffuse emission, they were able to deconvolve an astrophysical observation as well as decomposing it into point-like and diffuse emission components. While the method has worked well for large-scale diffuse emission, such as the Galactic diffuse gamma-ray emission, it seemed to fail for smaller extended emission structures such as morphological details of extended sources.

@@

    Some of the challenges of deblurring with CNN are due to the convolutional nature of the networks. In a single layer, linear approximation, the network would need to learn the inverse filter of the PSF, corresponding to the well known \textit{Wiener filter} solution of the problem. This requires either large convolution kernels or deep network architectures, which require in turn large amounts of training data. This challenge was described early by \cite{Li2014} who introduced a baseline architecture for a deep CNN for image deblurring.

@@

    \begin{figure*}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/model-illustration.pdf}
            \caption{
                Illustration of the forward model of the low counts imaging process. The flux represents the true unknown flux distribution in the sky. The flux is convolved with an image of the PSF, which represents an estimate of the resolution of the imaging instrument. The exposure consists of the superposition of multiple independent observations with different observation time and pointing position. The exposure weighted flux image is degraded by Poisson noise.
            }
            \label{fig:model-illustration}
            \script{model-illustration.py}
        \end{centering}
    \end{figure*}
    \section{Method}

    \subsection{Poisson Joint Likelihood}
    Our goal is to recover an image $\mathbf{x}$ from a set of multiple low counts observations. Most generally we assume our total dataset consists of $J$ individual observations of the same region of the sky, which we jointly model. The assumption is that the underlying, unknown \textit{true} emission image, we are interested in, does not change with time. Under this assumption these datasets can be for example:

@@

    Or even an arbitrary combination of the possibilities listed above. It could also be the combination of observations in different energy ranges, however in this case the assumption of the morphology not changing with energy would apply, which is rarely fulfilled for the astronomical data.

    For each individual observation $j$ the predicted counts can be modelled by \textit{forward folding} the unknown flux image $\mathbf{x}$ through the individual instrument response:
    %
    \begin{equation}
        \label{eq:model}
        \boldsymbol{\lambda}_j(\mathbf{x}) = \mathbf{E}_j \cdot \left(\mathrm{PSF}_j \circledast \mathbf{x}\right) + \mathbf{B}_j
    \end{equation}

@@

    Where $\mathbf{\lambda_j}$ depends on $\mathbf{x}$ as defined by Equation~\ref{eq:model}. By taking the logarithm and dropping the constant terms one can transform the product into a sum over pixels, which is also often called the \textit{Cash} \citep{Cash1979} fit statistics:
    %
    \begin{equation}
        \label{eq:cash}
        \mathcal{C}\left( \mathbf{D_j} | \boldsymbol{\lambda_j} \right) = \frac{1}{N}\sum_{i=1}^N \lambda_{j,i} - D_{j, i} \log{\lambda_{j,i}}
    \end{equation}

@@

    Using Maximum Likelihood Estimation (MLE) we could try to get an estimate of $\hat{\mathbf{x}}$ directly, using some optimization procedure. However, in general this represents an \textit{ill posed} inverse problem. Estimates for $\mathbf{x}$ which have similar likelihood values $\mathcal{L}(\hat{\mathbf{x}}_1)$ look very different from each other and might also look different compared to a given ground truth, if available. Both the convolution operation as well as the existence of noise lead to a loss of information in the data which cannot easily be recovered. The likelihood surface shows many local, close-by minima. Depending on the initial estimate for $\mathbf{x}$ and the minimization algorithm the MLE based estimate is likely to only find such a local minimum. This is fundamentally the reason for RL decomposing estimates into point sources, described in the previous section.

@@

    One convenient property of the G MM is the possibility to evaluate its likelihood in closed form. For a given single patch $\mathbf{x}_n$ it is given by:
    %
    \begin{equation}
        \label{eq:gmm}
        P_{GMM}(\mathbf{x}_n) = \sum_{k=1}^K \pi_k \mathcal{N}(\mathbf{x}_n| \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k^2)
    \end{equation}

@@

 \begin{figure}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/patches.pdf}
            \caption{
                Grid of overlapping quadratic patches of size $8\times8$~pixels on top of an example image of the size $128 \times 68$~pixels. The overlap size was chosen to be $2$ pixels, which results in 50 patches in total. The patches are shown translucent, such that the horizontal and vertical \enquote{stripes} result from two neighboring overlapping patches, while the small, even lighter squares result from four patches overlapping at the their corners. The bottom row shows the red example patch $\mathbf{x}_n$ on the left. The three other patches show the quantity $\Sigma_{k_{GMM}} \mathbf{x}_n$, which is the result of the patch represented in the \textit{Eigenbasis} of the covariance matrices of the 3 most likely GMM components $k_{GMM} = 105, 77, 17$. This example uses the GMM learned from the JWST Cas A image.
            }
            \label{fig:patches}
            \script{patches.py}
        \end{centering}
    \end{figure}

@@

    \subsubsection{Image Dynamic Range and Patch Normalisation}
    \label{ssec:patch-norm}
    Astronomical images typically show a much higher dynamic range compared to natural images. This is mostly due to the existence of point sources, which rarely occur in natural images. Astronomical point sources typically relate to compact bright objects, such as active galactic nuclei at high distances, or binary objects and stars at Galactic distances. In addition the visible matter in the universe is sparse such that astronomical images typically show sparse bright emission on a dark background.

    To normalise the patches \cite{Zoran2011} proposed to subtract the patch mean before learning the GMM. This leads to the GMM modeling the same morphological features at different levels of contrast as different components. While this proved to be beneficial in the context of reconstructing natural images it does not scale to images with high dynamic range, because every morphological feature would need to be present at many levels of intensity. While this could be handled by introducing a lot more components to the GMM, it seems unfeasible in practice due to the limited availability of training data and computational resources. For this reason we propose the following modified normalisation for the patches instead:
    %
    \begin{equation}
        \label{eq:patch-norm}
        \tilde{\mathbf{x}}_n = \frac{\mathbf{x}_n - \Bar{\mathbf{x}}_n}{\Bar{\mathbf{x}}_n}
    \end{equation}

@@

    \subsection{Jolideco GMM Library}
    \label{ssec:jolideco-gmm-library}
    In general the strength of the GMM prior is its high \textit{expressiveness}. This means the prior contains a large variety of structural shapes and has a lot of freedom to adapt to the data in the optimization process. On radio data \cite{Bouman2016} found their results to be mostly independent of the choice of data the GMM prior was learned on. However, in the low signal-to-noise regime we typically expect a higher dependence of the results on the choice of the data the GMM was learned on. The choice of the prior becomes more important and to avoid introducing a strong bias it should be adapted to the analysis scenario. For this reason and for purposes of testing we learned and provide a selection of GMMs (a \textit{Library}) to be used with the patch prior for download in a \textit{GitHub} repository\footnote{\url{https://github.com/jolideco/jolideco-gmm-library}}.

@@

    \subsubsection{JWST Cassiopeia A}
    As a specific prior for the analysis of SNRs we learned a GMM prior from data of the James Webb Space Telescope (JWST).
    The JWST is a large, infrared space telescope launched in late 2021. It delivers image data at very high spatial resolution as well as excellent signal to noise ratio. Both of these properties make JWST data an excellent choice for extracting patches for training a GMM. To limit the data reduction effort we relied on a promotional image of the SNR \textit{Cassiopeia A}\footnote{We downloaded the full resolution (4k)  PNG image from \url{https://webbtelescope.org/contents/media/images/2023/121/01GWQBBY77MHGFV3M3N63KDCEJ}}.
    We first converted the image to gray scale using a perceived brightness model \citep{scikit-image}. Then we split the image into $\approx 2$ million patches and learned a GMM with 128 mixture components. Again we relied on the patch normalisation described in Section~\ref{ssec:patch-norm}. Figure~\ref{fig:gmm-eigen-images} shows the weighted sum of the \textit{Eigenimages} of the covariance matrices of the first 32 components of the GMM. The patches show a large variety in morphological shapes, including edges, gradients, bright compact spots and more combinations thereof.

@@

    To improve the convergence rate of the optimization process we employ the following optimization strategy: the optimization strategy begins with the optimizer viewing $\log{\mathbf{x}}$, and the first step involves running a rough estimate using a uniform prior. This step is optional, but if chosen, the next step involves optimizing the parameters associated with the systematic error described in Eq.~\ref{eq:model-npred-calibration}. Following this, a full optimization is performed using the patch prior, which can be computationally expensive. The optimization algorithm used is ADAM \citep{Kingma2014}, a stochastic gradient descent method, without decay and a fixed step size value of $\mathrm{lr}=10^{-3}$ has been used.

@@

    \subsection{Results}
    \subsubsection{Comparing Methods}
    %
    \begin{figure*}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-scenarios.pdf}
            \caption{
                Comparison of different deconvolution methods and prior assumptions for different selected source scenarios, as described in Section~\ref{subsec:test-datasets}. The simulation of the data for this figure used a fixed background level of $\lambda_{Bkg} = \qty[mode = math]{0.01}{cts/pix}$, a uniform exposure and a Gaussian PSF of width $\sigma_{PSF} = \qty[mode = math]{2}{pix}$, corresponding to the \enquote{Chandra} instrument scenario. The leftmost column shows the simulated counts data and the second most column from the left the underlying ground truth to compare. The other columns show the reconstructed flux by the different methods and prior assumptions for \jolideco. To enhance weak structures the images use an $\arcsinh$ stretching with a scale parameter of $a=0.02$. The stretching is the same for all images. The methods are described in detail in Section~\ref{subsec:methods}. A more detailed representation of the results is available at \url{https://jolideco.github.io/jolideco-comparison}.
            }
            \label{fig:comparison-scenarios}
            \script{comparison-scenarios.py}
        \end{centering}
    \end{figure*}
    %
    First we compare the performance of methods and different prior assumptions on the different source scenarios. Figure \ref{fig:comparison-scenarios} shows the results for the selected subset of scenarios A1, B3, C3 and D4.

@@

    \textbf{Zoom B} shows the region around the central compact object of the SNR. While the emission is barely visible in the counts image it clearly sticks out above the background in the image reconstructed by \jolideco. The remaining visible extension in the point source is encoded in the JWST patch prior, which also shows a finite extension for point sources. We note that, as shown in Section~\ref{subsec:methods} for scenario A, the apparent size of the point source was smaller if the source was brighter. We estimated the position of point source by finding the brightest pixel in the map. We obtained $\alpha_{J2000}=01^h04^m02.75^s$ and $\delta_{J2000}=-72^d02^m00.19^s$ which is consistent with the value reported by \cite{Long2020}.

@@

    \subsubsection{Flux Profile}
    To get a more quantitative measurement of the improvement in angular resolution we compute a flux profile on the image of E0102 reconstructed by \jolideco. For this we chose a region perpendicular to the filament shown in Zoom A of Figure~\ref{fig:example-chandra}. To estimate the errors we use the bootstrapping method. For each individual observation we sample from the event list with replacement. For each realization of the set of observation we run \jolideco and compute a flux profile on the reconstructed image. This way we obtain a distribution of profiles, which we can use to compute an error band. For comparison we also compute the profiles and associated errors directly on the stacked bootstrapped images of counts. The result is shown in Figure~\ref{fig:chandra-e0102-zoom-a}.

@@

    Assuming that the PSF and the shape of peak of the profile are approximately Gaussian we expect the following relation between the width of the counts profile, the PSF and the flux profile:
    %
    \begin{equation}
        \omega_{J} = \sqrt{\omega_C^2 - \omega_{\mathrm{PSF}}^2}
    \end{equation}
    %
    With $w_{\mathrm{PSF}} \approx 0.6 \mathrm{arcsec}$ \footnote{See projected radial profiles of the on-axis PSF at \url{https://space.mit.edu/cxc/marx/tests/PSF.html}.  The 1$\sigma$ radius is $\approx$0.26~arcsec (corresponding to 39\% enclosed counts in 2-D), suggesting that FWHM$_{{\rm PSF}}\approx{2.355{\cdot}0.26=0.62}$~arcsec.}, we expect a FWHM after de-convolution of $\approx\omega_J=0.48$~arcsec. This is roughly consistent with the values we measured for the profile.

@@

    \subsection{\fermi Event Types}
     We apply the \jolideco method to data from the Fermi Large Area Telescope (LAT). The LAT is a pair conversion \gammaray telescope on board of the Fermi satellite. It operates in the energy range between \qty[mode = text]{30}{MeV} to  \qty[mode = text]{300}{GeV}. For this example we chose  a Galactic source with complex morphology: the supernova remnant Vela Junior. We use 14 years (from 2008 August 4 to 2023 August 4) of \endquote{P8R3\_SOURCE} photons with reconstructed energy in the range of \qty[mode = text]{10}{GeV} to  \qty[mode = text]{2}{TeV} and applied a zenith angle cut of \ang{105}. To preserve as much information as possible on the angular resolution of the instrument we split the data into the four event types \enquote{PSF0}, \enquote{PSF1}, \enquote{PSF2} and \enquote{PSF3}, as defined by the \fermi collaboration. We chose a pixel size of \qty[mode = text]{0.02}{$\deg$}, which corresponds to approximately 1/5th of the size of the PSF in the best event class. For the first part of the data reduction we use the standard \texttt{fermitools v2.2.0} \cite{}.

@@

    \paragraph{Asterism} This pattern is comprised of a diffuse Gaussian source ($\sigma=2$~pix) at the center of the image at pixel location (64,64), and four larger extended sources (Gaussians with $\sigma=4$~pix) with the same number of counts as the central source are set at each of the corners.  Four point sources with intensities decreasing by a factor $\frac{1}{3}$ are arranged around the central diffuse source to the top, left, right, and bottom.  Three brightness levels are considered, with counts in all the diffuse sources set at 
    $$\{100, 500, 1000\} \cdot \tau \,,$$
    where $\tau=1$ for \textit{Chandra}-like and $\tau=5$ for XMM-like simulations.  The diffuse sources are located at
    $$(x,y)_{{\rm diffuse}} = (22,22), (22,106), (106,106), (106,22) ~{\rm pix} \,,$$
    and the point sources at
    $$(x,y)_{{\rm point}} = (64,76), (44,64), (74,64), (64,44) \,,$$
    with counts
    $$\{1, \frac{1}{3}, \frac{1}{3^2}, \frac{1}{3^3} \} \times$$
    those in the diffuse sources respectively.  The pattern of point sources explores the effects of contamination by large extended sources, and the diffuse sources at the corners help to characterize the edge effects in deconvolutions.

>>>>

% Filler text
\usepackage{blindtext}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{showyourwork}
\usepackage[autostyle]{csquotes}
\usepackage{amssymb}
\let\tablenum\relaxgit 
\usepackage{siunitx}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\arcsinh}{arcsinh}
\newcommand{\chandra}{\textit{Chandra}~}
\newcommand{\xmm}{\textit{XMM}~}
\newcommand{\fermi}{\textit{Fermi}-LAT~}
\newcommand{\jolideco}{\textit{Jolideco}~}
\newcommand{\jolideconospace}{\textit{Jolideco}}
\newcommand{\aposteriori}{a~posteriori~}
\newcommand{\gammaray}{$\gamma$-ray\xspace}
\newcommand{\xray}{X-ray\xspace}
\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}\PackageWarning{TODO:}{#1!}}

@@

    \section{Introduction}
    The resolution of any physical or astronomical imaging process is inherently limited by the instrument or telescope used. Furthermore, the quality of the resulting image is affected by background or instrumental measurement noise and non-uniform exposure. This is especially true in the low signal-to-noise regime, where images are affected by Poisson noise. This includes images taken by \xray and \gammaray~telescopes but also medical images obtained with methods such as positron emission tomography or single-photon emission computed tomography. This low counts imaging process is illustrated in Figure~\ref{fig:model-illustration}.

    To maximise the scientific use of available data it is desirable to correct for the instrumental imprint on the data as well as reduce its noise level. The process of improving the resolution of an image in post-processing is often called \enquote{deblurring} or \enquote{deconvolution}, as it aims to invert the effect of the data being degraded by a forward convolution through the point spread function (PSF) of the instrument. In general this class of problems is categorised as \enquote{inverse problems}. If there is prior information available on the shape of the PSF, we typically refer to the process more precisely as \enquote{unblind deconvolution}. In case of astronomical observations the PSF can for example be estimated empirically from observation of known point sources, such as distant stars, or a model of the PSF can be obtained from simulation of the telescope or instrument.

    \begin{figure*}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/model-illustration.pdf}
            \caption{
                Illustration of the forward model of the low counts imaging process. The flux represents the true unknown flux distribution in the sky. The flux is convolved with an image of the PSF, which represents an estimate of the resolution of the imaging instrument. The exposure consists of the superposition of multiple independent observations with different observation time and pointing position. The exposure weighted flux image is degraded by Poisson noise.
            }
            \label{fig:model-illustration}
            \script{model-illustration.py}
        \end{centering}
    \end{figure*}

%\as{You may want to add a short para or just a sentance saying that X-ray and gamma-ray images are the Poisson type images containing low number of counts in each image pixel etc.. - this is to link to the next subsection which is specific to Poisson noise?}
% AD: this is mentioned already...

@@

    The standard baseline method for image deconvolution in the presence of Poisson noise was proposed independently by \cite{Richardson1972} and \cite{Lucy1974} (we will further refer to this method as \enquote{RL}). The method involves estimating the reconstructed image by an iterative procedure that converges to a maximum likelihood solution of the problem.

@@

    \cite{Selig2015} and \cite{Pumpe2018} approached the problem of deconvolution of Poisson noise data implicitly using the framework of information field theory. Based on physical prior assumptions on the spatial distribution of point sources and local correlation structures of diffuse emission, they were able to deconvolve \fermi observations as well as decomposing it into point-like and diffuse emission components. While their method has worked well for large-scale diffuse emission, such as the Galactic diffuse gamma-ray emission, it showed artifacts for smaller extended emission structures such as morphological details of extended sources in the Galactic Plane.

@@

    Some of the challenges of deblurring with CNN are due to the convolutional nature of the networks. In a single layer, linear approximation, the network would need to learn the inverse filter of the PSF, corresponding to the well known \enquote{Wiener Filter} solution of the problem. This requires either large convolution kernels or deep network architectures, which require in turn large amounts of training data. This challenge was described early by \cite{Li2014} who introduced a baseline architecture for a deep CNN for image deblurring.

@@

    \section{Method}

    \subsection{Poisson Joint Likelihood}
    Our goal is to recover an image $\mathbf{x}$ from a set of multiple low counts observations. Most generally we assume our total dataset consists of $J$ individual observations of the same region of the sky, which we jointly model. The assumption is that the underlying, unknown \enquote{true emission} image, we are interested in, does not change with time. Under this assumption these datasets can be for example:

@@

    Or even an arbitrary combination of the possibilities listed above. It could also be the combination of observations in different energy ranges, however in this case the assumption of the morphology not changing with energy would apply, which is rarely fulfilled for astronomical objects.

    For each individual observation $j$ the predicted counts can be modelled by forward folding the unknown flux image $\mathbf{x}$ through the individual instrument response:
    %
    \begin{equation}
        \label{eq:model}
        \boldsymbol{\lambda}_j(\mathbf{x}) = \mathbf{E}_j \cdot \left(\mathrm{PSF}_j \circledast \mathbf{x}\right) + \mathbf{B}_j
    \end{equation}

@@

    Where $\mathbf{\lambda_j}$ depends on $\mathbf{x}$ as defined by Equation~\ref{eq:model}. By taking the logarithm and dropping the constant terms one can transform the product into a sum over pixels, which is also often called the \enquote{Cash} \citep{Cash1979} fit statistics:
    %
    \begin{equation}
        \label{eq:cash}
        \mathcal{C}\left( \mathbf{D_j} | \boldsymbol{\lambda_j} \right) = \frac{1}{N}\sum_{i=1}^N \lambda_{j,i} - D_{j, i} \log{\lambda_{j,i}}
    \end{equation}

@@

    Using Maximum Likelihood Estimation (MLE) we could try to get an estimate of $\hat{\mathbf{x}}$ directly, using some optimization procedure. However, in general this represents an \enquote{ill posed} inverse problem. Estimates for $\mathbf{x}$ which have similar likelihood values $\mathcal{L}(\hat{\mathbf{x}}_1)$ look very different from each other and might also look different compared to a given ground truth, if available. Both the convolution operation as well as the existence of noise lead to a loss of information in the data which cannot easily be recovered. The likelihood surface shows many local, close-by minima. Depending on the initial estimate for $\mathbf{x}$ and the minimization algorithm the MLE based estimate is likely to only find such a local minimum. This is fundamentally the reason for RL decomposing estimates into point sources, described in the previous section.

@@

    One convenient property of the GMM is the possibility to evaluate its likelihood in closed form. For a given single patch $\mathbf{x}_n$ it is given by:
    %
    \begin{equation}
        \label{eq:gmm}
        P_{GMM}(\mathbf{x}_n) = \sum_{k=1}^K \pi_k \mathcal{N}(\mathbf{x}_n| \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k^2)
    \end{equation}

@@

 \begin{figure}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/patches.pdf}
            \caption{
                Grid of overlapping quadratic patches of size $8\times8$~pixels on top of an example image of the size $128 \times 68$~pixels. The overlap size was chosen to be $2$ pixels, which results in 50 patches in total. The patches are shown translucent, such that the horizontal and vertical \enquote{stripes} result from two neighboring overlapping patches, while the small, even lighter squares result from four patches overlapping at the their corners. The bottom row shows the red example patch $\mathbf{x}_n$ on the left. The three other patches show the quantity $\Sigma_{k_{GMM}} \mathbf{x}_n$, which is the result of the patch represented in the Eigenbasis of the covariance matrices of the 3 most likely GMM components $k_{GMM} = 105, 77, 17$. This example uses the GMM learned from the JWST Cas A image.
            }
            \label{fig:patches}
            \script{patches.py}
        \end{centering}
    \end{figure}

@@

    \subsubsection{Image Dynamic Range and Patch Normalisation}
    \label{ssec:patch-norm}
    Astronomical images typically show a much higher dynamic range compared to natural images.
    % This is mostly due to the existence of point sources, which rarely occur in natural images. Astronomical point sources typically relate to compact bright objects, such as unresolved active galactic nuclei at high distances, or binary objects and stars at Galactic distances. In addition the visible matter in the universe is sparse such that astronomical images typically show sparse bright emission on a dark background. 
    To normalise the patches \cite{Zoran2011} proposed to subtract the patch mean before learning the GMM. This leads to the GMM modeling the same morphological features at different levels of contrast as different components. While this proved to be beneficial in the context of reconstructing natural images it does not scale to images with high dynamic range, because every morphological feature would need to be present at many levels of intensity. While this could be handled by introducing a lot more components to the GMM, it seems unfeasible in practice due to the limited availability of training data and computational resources. For this reason we propose the following modified normalisation for the patches instead:
    %
    \begin{equation}
        \label{eq:patch-norm}
        \tilde{\mathbf{x}}_n = \frac{\mathbf{x}_n - \Bar{\mathbf{x}}_n}{\Bar{\mathbf{x}}_n}
    \end{equation}

@@

    \subsection{Jolideco GMM Library}
    \label{ssec:jolideco-gmm-library}
    In general the strength of the GMM prior is its high \enquote{expressiveness}. This means the prior contains a large variety of structural shapes and has a lot of freedom to adapt to the data in the optimization process. On radio data \cite{Bouman2016} found their results to be mostly independent of the choice of data the GMM prior was learned on. However, in the low signal-to-noise regime we typically expect a higher dependence of the results on the choice of the data the GMM was learned on. The choice of the prior becomes more important and to avoid introducing a strong bias it should be adapted to the analysis scenario. For this reason and for purposes of testing we learned and provide a selection of GMMs (a library) to be used with the patch prior for download in a \textit{GitHub} repository\footnote{\url{https://github.com/jolideco/jolideco-gmm-library}}.

@@

    \subsubsection{JWST Cassiopeia A}
    As a specific prior for the analysis of SNRs we learned a GMM prior from data of the James Webb Space Telescope (JWST).
    The JWST is a large, infrared space telescope launched in late 2021. It delivers image data at very high spatial resolution as well as excellent signal to noise ratio. Both of these properties make JWST data an excellent choice for extracting patches for training a GMM. To limit the data reduction effort we relied on a promotional image of the SNR \textit{Cassiopeia A}\footnote{We downloaded the full resolution (4k)  PNG image from \url{https://webbtelescope.org/contents/media/images/2023/121/01GWQBBY77MHGFV3M3N63KDCEJ}}.
    We first converted the image to gray scale using a perceived brightness model from the Scikit-Image package \citep{scikit-image}. Then we split the image into $\approx 2$ million patches and learned a GMM with 128 mixture components. Again we relied on the patch normalisation described in Section~\ref{ssec:patch-norm}. Figure~\ref{fig:gmm-eigen-images} shows the weighted sum of the \textit{Eigenimages} of the covariance matrices of the first 32 components of the GMM. The patches show a large variety in morphological shapes, including edges, gradients, bright compact spots and more combinations thereof.

@@

    To improve the convergence rate of the optimization process we employ the following optimization strategy: the optimization strategy begins with the optimizer viewing $\log{\mathbf{x}}$, and the first step involves running a rough estimate using a uniform prior. This step is optional, but if chosen, the next step involves optimizing the parameters associated with the systematic error described in Eq.~\ref{eq:model-npred-calibration}. Following this, a full optimization is performed using the patch prior, which can be computationally expensive. The optimization algorithm used is ADAM\citep{Kingma2014}, a stochastic gradient descent method, without decay and a fixed step size value of $\mathrm{lr}=10^{-3}$ has been used.

@@

    \subsection{Results}
    \subsubsection{Comparing Methods}
    %
    \begin{figure*}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-scenarios.pdf}
            \caption{
                Comparison of different deconvolution methods and prior assumptions for different selected source scenarios, as described in Section~\ref{subsec:test-datasets}. The simulation of the data for this figure used a fixed background level of $\lambda_{Bkg} = \qty[mode = math]{0.01}{cts/pix}$, a uniform exposure and a Gaussian PSF of width $\sigma_{PSF} = \qty[mode = math]{2}{pix}$, corresponding to the \enquote{Chandra} instrument scenario. The leftmost column shows the simulated counts data and the second most column from the left the underlying ground truth to compare. The other columns show the reconstructed flux by the different methods and prior assumptions for \jolideconospace. To enhance weak structures the images use an $\arcsinh$ stretching with a scale parameter of $a=0.02$. The stretching is the same for all images. The methods are described in detail in Section~\ref{subsec:methods}. A more detailed representation of the results is available at \url{https://jolideco.github.io/jolideco-comparison}.
            }
            \label{fig:comparison-scenarios}
            \script{comparison-scenarios.py}
        \end{centering}
    \end{figure*}
    %
    First we compare the performance of methods and different prior assumptions on the different source scenarios. Figure \ref{fig:comparison-scenarios} shows the results for the selected subset of scenarios A1, B3, C3 and D4.

@@

    \textbf{Zoom B} shows the region around the central compact object of the SNR. While the emission is barely visible in the counts image it clearly sticks out above the background in the image reconstructed by \jolideconospace. The remaining visible extension in the point source is encoded in the JWST patch prior, which also shows a finite extension for point sources. We note that, as shown in Section~\ref{subsec:methods} for scenario A, the apparent size of the point source was smaller if the source was brighter. We estimated the position of point source by finding the brightest pixel in the map. We obtained $\alpha_{J2000}=01^h04^m02.75^s$ and $\delta_{J2000}=-72^d02^m00.19^s$ which is consistent with the value reported by \cite{Long2020}.

@@

    \subsubsection{Flux Profile}
    To get a more quantitative measurement of the improvement in angular resolution we compute a flux profile on the image of E0102 reconstructed by \jolideconospace. For this we chose a region perpendicular to the filament shown in Zoom A of Figure~\ref{fig:example-chandra}. To estimate the errors we use the bootstrapping method. For each individual observation we sample from the event list with replacement. For each realization of the set of observation we run \jolideco and compute a flux profile on the reconstructed image. This way we obtain a distribution of profiles, which we can use to compute an error band. For comparison we also compute the profiles and associated errors directly on the stacked bootstrapped images of counts. The result is shown in Figure~\ref{fig:chandra-e0102-zoom-a}.

@@

    Assuming that the PSF and the shape of peak of the profile are approximately Gaussian we expect the following relation between the width of the counts profile, the PSF and the flux profile:
    %
    \begin{equation}
        \omega_{J} = \sqrt{\omega_C^2 - \omega_{\mathrm{PSF}}^2}
    \end{equation}
    %
    With $\omega_{\mathrm{PSF}}\approx0.6$~arcsec\footnote{See projected radial profiles of the on-axis PSF at \url{https://space.mit.edu/cxc/marx/tests/PSF.html}.  The 1$\sigma$ radius is $\approx$0.26~arcsec (corresponding to 39\% enclosed counts in 2-D), suggesting that FWHM$_{{\rm PSF}}\approx{2.355{\cdot}0.26=0.62}$~arcsec.}, we expect a FWHM after de-convolution of $\omega_J\approx0.48$~arcsec. This is roughly consistent with the values we measured for the profile.

@@

    \subsection{\fermi Event Types}
     We apply the \jolideco method to data from the Fermi Large Area Telescope (LAT) \citep{Atwood2009}. The LAT is a pair conversion \gammaray telescope on board of the Fermi satellite. It operates in the energy range between \qty[mode = text]{30}{MeV} to  \qty[mode = text]{300}{GeV}. For this example we chose  a Galactic source with complex morphology: the supernova remnant Vela Junior. We use 14 years (from 2008 August 4 to 2023 August 4) of \enquote{P8R3\_SOURCE} photons with reconstructed energy in the range of \qty[mode = text]{10}{GeV} to \qty[mode = text]{2}{TeV} and applied a zenith angle cut of \ang{105}. To preserve as much information as possible on the angular resolution of the instrument we split the data into the four event types \enquote{PSF0}, \enquote{PSF1}, \enquote{PSF2} and \enquote{PSF3}, as defined by the \fermi \enquote{Pass8} data release\footnote{\url{https://fermi.gsfc.nasa.gov/ssc/data/analysis/documentation/Pass8_usage.html}}. We chose a pixel size of \qty[mode = text]{0.02}{$\deg$}, which corresponds to approximately 1/5 of the size of the PSF in the best event class. For the first part of the data reduction we use the standard \texttt{fermitools v2.2.0} \cite{Fermitools2019}.

@@

    \paragraph{Asterism} This pattern is comprised of a diffuse Gaussian source ($\sigma=2$~pix) at the center of the image at pixel location (64,64), and four larger extended sources (Gaussians with $\sigma=4$~pix) with the same number of counts as the central source are set at each of the corners.  Four point sources with intensities decreasing by a factor $\frac{1}{3}$ are arranged around the central diffuse source to the top, left, right, and bottom.  Three brightness levels are considered, with counts in all the diffuse sources set at 
    $$\{100, 500, 1000\} \cdot \tau \,,$$
    where $\tau=1$ for \chandra-like and $\tau=5$ for \xmm-like simulations.  The diffuse sources are located at
    $$(x,y)_{{\rm diffuse}} = (22,22), (22,106), (106,106), (106,22) ~{\rm pix} \,,$$
    and the point sources at
    $$(x,y)_{{\rm point}} = (64,76), (44,64), (74,64), (64,44) \,,$$
    with counts
    $$\{1, \frac{1}{3}, \frac{1}{3^2}, \frac{1}{3^3} \} \times$$
    those in the diffuse sources respectively.  The pattern of point sources explores the effects of contamination by large extended sources, and the diffuse sources at the corners help to characterize the edge effects in deconvolutions.
